---
title: "Introduction to lctmr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to lctmr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## What is Latent Class Trajectory Modeling?

Latent Class Trajectory Modeling (LCTM) identifies hidden subgroups (latent classes) of individuals who follow similar patterns of change over time. Unlike standard growth curve models that assume everyone follows the same trajectory, LCTM discovers distinct subgroups with different growth patterns.

For example, in pediatric growth data, LCTM might discover:

- **Class 1**: Children with consistently high weight
- **Class 2**: Children with rapid early growth that plateaus
- **Class 3**: Children with slow but steady growth
- **Class 4**: Children with declining growth

## Getting Started

```{r load-package}
library(lctmr)
```

### Sample Data

The package includes a sample dataset of pediatric growth measurements from the Mazira study:

```{r sample-data}
data(sample_growth)
head(sample_growth)

# Data structure
str(sample_growth)

# Number of children and observations
cat("Children:", length(unique(sample_growth$childid)), "\n")
cat("Total observations:", nrow(sample_growth), "\n")
```

## Two Ways to Use lctmr

### Option A: Investigator-Driven Workflow (Recommended)

This is the recommended approach, aligned with Lennon et al. (2018). The investigator examines diagnostic plots to make informed decisions about model specification before automated refinement.

```{r investigator-example, eval = FALSE}
library(lctmr)
data(sample_growth)

# Step 1: Clean data (optional but recommended)
cleaned <- lctm_clean(
  sample_growth,
  outcome = "weight_raw",
  time_var = "anthroage",
  id_var = "childid"
)

# Step 2: Fit initial model and examine diagnostics
initial <- lctm_initial(
  cleaned,
  k = 3,
  degree = 2,
  save_pdf = "initial_diagnostics.pdf"
)

# Review: spaghetti, loess, and residual plots guide your decisions
plot(initial)

# Step 3: Refine based on residual patterns
# e.g., if residuals show a diagonal trend, add a linear random slope:
result <- lctm_refine(
  initial,
  random = ~ 1 + anthroage,
  k_range = 2:5
)

summary(result)
plot(result)
```

### Splines as an alternative trajectory form

When residual diagnostics show non-polynomial patterns (e.g., a distinct change of slope around 6–12 months), specify natural-spline `knots` instead of a polynomial `degree`:

```{r splines-example, eval = FALSE}
initial_sp <- lctm_initial(cleaned, k = 3, knots = c(6, 12))
result_sp  <- lctm_refine(initial_sp, knots = c(6, 12), k_range = 2:5)
```

`lctm_refine()` carries the same `knots` through to all candidate models for a like-for-like comparison.

The `spline_degree` argument sets the polynomial order of the spline pieces. The
default `3` builds a natural cubic spline (`splines::ns()`); `2` builds a
quadratic B-spline (`splines::bs(degree = 2)`), which uses one fewer parameter
per knot and tends to estimate more stably in a multi-class mixture — useful
when a cubic spline is over-flexible or fails to converge:

```{r splines-quadratic, eval = FALSE}
initial_q <- lctm_initial(cleaned, k = 3, knots = c(6, 12), spline_degree = 2)
result_q  <- lctm_refine(initial_q, knots = c(6, 12), spline_degree = 2, k_range = 2:5)
```

Like `knots`, `spline_degree` is inherited by `lctm_refine()` from the initial
model unless you override it.

## Understanding Model Types

lctmr supports two model types, evaluated during refinement:

| Model | Name | Variance Structure | When to Use |
|-------|------|-------------------|-------------|
| A | Common variance | Same across classes | Simpler, homogeneous classes |
| B | Proportional variance | Proportional to class size | More flexible, often fits better |

Model B is tried first by default, as it allows classes to have different levels of within-class variability.

## Understanding Adequacy Metrics

Three metrics determine if a model adequately separates classes:

| Metric | Threshold | Meaning |
|--------|-----------|---------|
| **APPA** | > 0.70 | Average Posterior Probability of Assignment - how confident are class assignments? |
| **OCC** | > 5.0 | Odds of Correct Classification - odds that assignment is correct vs. random |
| **Entropy** | > 0.5 | Relative Entropy - overall class separation quality |

A model passes adequacy if **all three** metrics meet their thresholds.

## What's Next?

- See `vignette("workflow")` for detailed explanation of the 4-step methodology
- See `?lctm_initial` and `?lctm_refine` for the investigator-driven workflow
- See `?lctm_plot_trajectories` for visualization options (legend includes per-class n and %)
