---
title: "The 5-Step LCTM Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The 5-Step LCTM Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

This vignette explains the complete 5-step workflow for Latent Class Trajectory Modeling. Understanding this workflow helps you make informed decisions when the automated approach needs adjustment.

```{r load}
library(lctmr)
data(sample_growth)
```

## Overview

The LCTM workflow is **iterative**. Steps 3-4 may be repeated multiple times until you find a model that passes adequacy checks.
```
Step 1: Determine random effects structure (once)
           ↓
Step 2: Compare BIC across K values (once)
           ↓
    ┌─→ Step 3: Fit candidate model
    │      ↓
    │   Step 4: Check adequacy
    │      ↓
    │   Pass? ──No──→ Try different K or model type ─┐
    │      │                                         │
    │     Yes                                        │
    │      ↓                                         │
    │   Step 5: Visualize                            │
    │                                                │
    └────────────────────────────────────────────────┘
```

## Step 1: Determine Random Effects Structure

The random effects structure controls model flexibility. This is determined once at the start based on your data characteristics.

**Question**: Does your outcome need random intercept only, random slope, or random quadratic term?

For most growth data, a **quadratic random effects structure** works well because growth often follows curved patterns. This is the default in lctmr.

```{r step1, eval = FALSE}
# The setup function fits a preliminary model to assess structure
setup <- lctm_setup(
  sample_growth,
  outcome = "weight_raw",
  time_var = "anthroage",
  id_var = "childid",
  preliminary_k = 3,  # Use 3 classes for preliminary assessment
  degree = 2          # 2 = quadratic (default), 1 = linear
)

# Check the residual analysis
setup$random_effects_check
```

**Decision guide:**

| Pattern in Residuals | Recommended Structure |
|---------------------|----------------------|
| Random scatter | Current structure is adequate |
| Funnel shape over time | May need different variance structure |
| Curved pattern | May need higher degree polynomial |

## Step 2: Compare BIC Across K Values

BIC (Bayesian Information Criterion) helps identify the optimal number of classes. **Lower BIC = better fit** (accounting for model complexity).

```{r step2, eval = FALSE}
# BIC comparison is done automatically in lctm_setup()
setup <- lctm_setup(
  sample_growth,
  outcome = "weight_raw",
  time_var = "anthroage",
  id_var = "childid",
  k_range = 1:7  # Test 1 to 7 classes
)

# View the BIC table
setup$bic_table

# See ranked recommendations
setup$k_ranking
```

**Interpreting BIC results:**

- Start with the K that has the lowest BIC
- If that model fails adequacy, try the next-best K
- Very similar BIC values (within ~10) suggest comparable fit
- Consider parsimony: prefer fewer classes if BIC is similar

## Step 3: Fit Candidate Models

Now fit models with your candidate K values. Test both Model E and Model F.

```{r step3, eval = FALSE}
# Model F (proportional variance) - usually fits better
model_f <- lctm_fit(setup, k = 4, model = "F")

# Model E (common variance) - simpler
model_e <- lctm_fit(setup, k = 4, model = "E")

# View model summary
print(model_f)
```

### Model E vs Model F

| Aspect | Model E | Model F |
|--------|---------|---------|
| Variance | Same for all classes | Proportional to class size |
| Flexibility | Less | More |
| Parameters | Fewer | More |
| When to use | Homogeneous classes | Heterogeneous classes |

**Recommendation**: Start with Model F. If it fails adequacy or doesn't converge, try Model E.

### Advanced Options

```{r step3-advanced, eval = FALSE}
# Use natural splines for complex trajectories
model_spline <- lctm_fit(setup, k = 4, model = "F",
                         use_splines = TRUE, spline_df = 3)

# Use linear instead of quadratic
model_linear <- lctm_fit(setup, k = 4, model = "F", linear = TRUE)
```

## Step 4: Check Model Adequacy

A model must pass **all three** adequacy criteria to be considered valid:

```{r step4, eval = FALSE}
adequacy <- lctm_adequacy(model_f)
print(adequacy)
```

### The Three Metrics

#### APPA (Average Posterior Probability of Assignment)

- **Threshold**: > 0.70
- **Meaning**: On average, how confident are we about class assignments?
- **Calculation**: For each class, average the posterior probability among members assigned to that class

```{r appa-explain, eval = FALSE}
# APPA for each class
adequacy$appa
# Example: c(0.85, 0.78, 0.82, 0.91) - all classes > 0.70 = PASS
```

#### OCC (Odds of Correct Classification)

- **Threshold**: > 5.0
- **Meaning**: What are the odds that a class assignment is correct vs. random?
- **Interpretation**: OCC of 10 means correct classification is 10x more likely than random

```{r occ-explain, eval = FALSE}
# OCC for each class
adequacy$occ
# Example: c(12.3, 8.5, 9.1, 15.2) - all classes > 5.0 = PASS
```

#### Relative Entropy

- **Threshold**: > 0.5
- **Meaning**: Overall measure of class separation (0 = no separation, 1 = perfect separation)
- **Single value**: Unlike APPA and OCC, entropy is one number for the whole model

```{r entropy-explain, eval = FALSE}
# Relative entropy
adequacy$entropy
# Example: 0.72 - above 0.5 = PASS
```

### What If Adequacy Fails?

```{r adequacy-fail, eval = FALSE}
adequacy <- lctm_adequacy(model_f)

if (!adequacy$overall_pass) {
  # Strategy 1: Try different K
  model_k3 <- lctm_fit(setup, k = 3, model = "F")
  adequacy_k3 <- lctm_adequacy(model_k3)

  # Strategy 2: Try different model type
  model_e <- lctm_fit(setup, k = 4, model = "E")
  adequacy_e <- lctm_adequacy(model_e)

  # Strategy 3: Try splines for complex patterns
  model_spline <- lctm_fit(setup, k = 4, model = "F", use_splines = TRUE)
  adequacy_spline <- lctm_adequacy(model_spline)

  # Strategy 4: Try linear if quadratic is overfitting
  model_linear <- lctm_fit(setup, k = 4, model = "F", linear = TRUE)
  adequacy_linear <- lctm_adequacy(model_linear)
}
```

**Troubleshooting guide:**

| Problem | Likely Cause | Solution |
|---------|--------------|----------|
| Low APPA in one class | Poorly defined class | Try fewer K |
| Low OCC everywhere | Classes overlap too much | Try fewer K or different model |
| Low entropy | Poor overall separation | Try fewer K |
| Model won't converge | Too many parameters | Try Model E or fewer K |

## Step 5: Visualize Results

Once you have a model that passes adequacy, visualize the results.

### Mean Trajectories

```{r step5-mean, eval = FALSE}
# Mean trajectory for each class with confidence intervals
lctm_plot_trajectories(model_f, type = "mean", ci = TRUE)
```

This plot shows:

- One line per class showing the average trajectory
- Shaded regions for 95% confidence intervals
- Class proportions in the legend

### Spaghetti Plots

```{r step5-spaghetti, eval = FALSE}
# Individual trajectories colored by class assignment
lctm_plot_trajectories(model_f, type = "spaghetti")
```

This plot shows:

- Every individual's trajectory as a thin line
- Colors indicate assigned class
- Helps assess within-class heterogeneity

### Combined View

```{r step5-both, eval = FALSE}
# Both mean and individual trajectories
lctm_plot_trajectories(model_f, type = "both")
```

### Customization

```{r step5-custom, eval = FALSE}
# Custom colors
lctm_plot_trajectories(model_f, type = "mean",
                       colors = c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3"))

# Custom time range
lctm_plot_trajectories(model_f, type = "mean",
                       time_range = c(0, 12))
```

## Complete Example

```{r complete-example, eval = FALSE}
library(lctmr)
data(sample_growth)

# Steps 1-2: Setup with BIC comparison
setup <- lctm_setup(
  sample_growth,
  outcome = "weight_raw",
  time_var = "anthroage",
  id_var = "childid"
)
print(setup$bic_table)

# Step 3-4: Iterative fitting
# Try K=4 first (assuming it had best BIC)
model <- lctm_fit(setup, k = 4, model = "F")
adequacy <- lctm_adequacy(model)
print(adequacy)

# If failed, try K=3
if (!adequacy$overall_pass) {
  model <- lctm_fit(setup, k = 3, model = "F")
  adequacy <- lctm_adequacy(model)
  print(adequacy)
}

# Step 5: Visualize final model
lctm_plot_trajectories(model, type = "mean", ci = TRUE)
lctm_plot_trajectories(model, type = "spaghetti")

# Extract class assignments for further analysis
assignments <- model$class_assignments
head(assignments)
```

## Summary

| Step | Function | Output |
|------|----------|--------|
| 1-2 | `lctm_setup()` | BIC table, K rankings |
| 3 | `lctm_fit()` | Fitted model object |
| 4 | `lctm_adequacy()` | APPA, OCC, entropy with pass/fail |
| 5 | `lctm_plot_trajectories()` | ggplot visualizations |

Remember: LCTM is iterative. Don't be discouraged if your first model doesn't pass adequacy. The workflow is designed for exploration and refinement.
