---
title: "The LCTM Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The LCTM Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

This vignette explains the investigator-driven workflow for Latent Class Trajectory Modeling, adapted from Lennon et al. (2018) and Gee (2014). The key idea is that the investigator examines diagnostic plots to make informed decisions about model specification before automated refinement.

```{r load}
library(lctmr)
data(sample_growth)
```

## Overview

The LCTM workflow has four steps:

```
Step 1: Prepare and clean longitudinal data
           |
Step 2: Visualize growth + fit initial model (no random effects)
           |
        Investigator reviews residual patterns
        and decides random effects structure
           |
Step 3: Automated refinement (BIC search + Model A/B + adequacy)
           |
Step 4: Interpret and display results
```

## Step 1: Prepare Longitudinal Data

Data cleaning identifies biologically implausible values using population-level and individual-level criteria.

```{r step1, eval = FALSE}
cleaned <- lctm_clean(
  sample_growth,
  outcome = "weight_raw",
  time_var = "anthroage",
  id_var = "childid",
  type = "weight",           # Apply weight-specific z-score criteria
  standards = "WHO",          # Use WHO growth chart thresholds
  zscore_col = "waz",         # Pre-computed WAZ column
  check_decrease = FALSE      # Only for height/HC data
)

print(cleaned)
```

**What `lctm_clean()` checks:**

| Criterion | Type | Details |
|-----------|------|---------|
| Missing values | General | Removes rows with NA in key variables |
| Z-score outliers | Population-level | WHO: WAZ < -6 or > 5; LAZ < -6 or > 6 |
| Z-score outliers | Population-level | CDC: WeightMZ < -5 or > 5; HeightMZ < -5 or > 3 |
| Birth weight | Population-level | Removes <= 500g or >= 5000g |
| Height/HC decrease | Individual-level | Flags decreases > 3cm between visits |
| Insufficient data | General | Removes subjects with < 2 observations |

This step is optional. You can pass a raw data frame directly to `lctm_initial()` if your data is already clean.

## Step 2: Visualize and Fit Initial Model

This step produces diagnostic plots and fits an initial model **without random effects**. The purpose is to examine residual patterns that guide the random effects structure for refinement.

```{r step2, eval = FALSE}
initial <- lctm_initial(
  cleaned,
  k = 3,                # Plausible number of classes
  degree = 2,           # 2 = quadratic (default), 1 = linear, 3 = cubic
  sex_var = NULL,        # Optional: column name for sex stratification
  save_pdf = "initial_diagnostics.pdf"
)

# View all diagnostic plots
plot(initial)

# Or view specific plots
plot(initial, which = "spaghetti")
plot(initial, which = "loess")
plot(initial, which = "residuals")  # Key diagnostic
plot(initial, which = "guide")      # Reference patterns
```

### Interpreting Residual Patterns

The residual plot is the key diagnostic. Compare the standardized residual patterns in each class against the reference guide:

| Pattern in Residuals | Random Effects Structure |
|---------------------|------------------------|
| Horizontal band | Random intercept only: `random = ~ 1` |
| Diagonal/linear trend | Add linear random slope: `random = ~ 1 + time_var` |
| Curved pattern | Add quadratic random effect: `random = ~ 1 + time_var + I(time_var^2)` |
| S-shaped pattern | Consider cubic terms or splines |

The initial model also produces spaghetti and loess plots. The loess plot helps characterize the functional form of growth (linear, quadratic, cubic) and identify potential inflection points that may require knots.

### Using Knots for Complex Trajectories

If the loess plot reveals inflection points, you can use natural splines with knots:

```{r step2-knots, eval = FALSE}
# Place knots at identified inflection points
initial_knots <- lctm_initial(
  cleaned,
  k = 3,
  degree = 1,            # Must be 1 when using knots
  knots = c(3, 6, 9)     # Knot positions (e.g., ages in months)
)
```

## Step 3: Automated Refinement

Based on the residual patterns from Step 2, specify the random effects structure and let `lctm_refine()` automate the search across K values and model types.

```{r step3, eval = FALSE}
# Example: residuals showed a diagonal trend -> linear random slope
result <- lctm_refine(
  initial,
  random = ~ 1 + anthroage,          # Based on residual interpretation
  k_range = 2:5,                      # K values to test
  models = c("B", "A"),               # Model B (proportional) tried first
  adequacy_thresholds = list(
    appa = 0.70,
    occ = 5.0,
    entropy = 0.5
  )
)

summary(result)
```

### What `lctm_refine()` Does Automatically

For each K value (ranked by BIC, lowest first):

1. Fits **Model B** (proportional variance) - allows classes to differ in variance
2. Checks adequacy (APPA, OCC, entropy)
3. If Model B fails, fits **Model A** (common variance)
4. Checks adequacy again
5. Stops at first model passing all criteria

If no model passes, the function reports this and suggests reconsidering the initial model specification.

### Model A vs Model B

| Aspect | Model A | Model B |
|--------|---------|---------|
| Variance | Same for all classes | Proportional to class size |
| Flexibility | Less | More |
| Parameters | Fewer | More |
| When to use | Homogeneous classes | Heterogeneous classes |

### Additional Options

```{r step3-options, eval = FALSE}
# Include covariates (added to fixed effects only)
result <- lctm_refine(
  initial,
  random = ~ 1 + anthroage,
  covariates = c("sex", "birth_weight"),
  k_range = 2:5
)

# Use simple starting values (more efficient for large datasets)
result <- lctm_refine(
  initial,
  random = ~ 1 + anthroage,
  start_simple = TRUE,
  k_range = 2:5
)

# Save result plots to PDF
result <- lctm_refine(
  initial,
  random = ~ 1 + anthroage,
  save_pdf = "refine_results.pdf"
)
```

## Step 4: Interpret and Display Results

Once you have a model that passes adequacy, examine the results.

### Adequacy Metrics

```{r step4-adequacy, eval = FALSE}
# Adequacy is stored in the result
print(result$adequacy)
```

### The Three Metrics

#### APPA (Average Posterior Probability of Assignment)

- **Threshold**: > 0.70
- **Meaning**: On average, how confident are we about class assignments?
- **Per-class**: Each class must individually meet this threshold

#### OCC (Odds of Correct Classification)

- **Threshold**: > 5.0
- **Meaning**: What are the odds that a class assignment is correct vs. random?
- **Interpretation**: OCC of 10 means correct classification is 10x more likely than random

#### Relative Entropy

- **Threshold**: > 0.5
- **Meaning**: Overall measure of class separation (0 = no separation, 1 = perfect)
- **Single value** for the whole model, unlike APPA and OCC which are per-class

### Trajectory Plots

```{r step4-mean, eval = FALSE}
# Mean trajectory for each class with confidence intervals
plot(result)

# Or call the plot function directly for more control
lctm_plot_trajectories(result$best_model, type = "mean", ci = TRUE)

# Individual trajectories colored by class assignment
lctm_plot_trajectories(result$best_model, type = "spaghetti")

# Both views
lctm_plot_trajectories(result$best_model, type = "both")
```

### Residual Diagnostics

```{r step4-residuals, eval = FALSE}
# Check residuals of the final model
lctm_plot_residuals(result$best_model, standardized = TRUE, by_class = TRUE)
```

### Class Assignments

```{r step4-assignments, eval = FALSE}
# Extract class assignments for further analysis
assignments <- result$class_assignments
head(assignments)

# Class proportions
table(assignments$class) / nrow(assignments)
```

### Customization

```{r step4-custom, eval = FALSE}
# Custom colors
lctm_plot_trajectories(result$best_model, type = "mean",
                       colors = c("#E41A1C", "#377EB8", "#4DAF4A"))

# Custom time range
lctm_plot_trajectories(result$best_model, type = "mean",
                       time_range = c(0, 12))
```

### What If No Model Passes Adequacy?

If `lctm_refine()` reports that no model met all criteria:

1. **Reconsider the initial model** - try a different trajectory shape or degree
2. **Adjust knot placement** - inflection points may be at different positions
3. **Try a different random effects structure** - the residual patterns may suggest an alternative
4. **Expand K range** - try more or fewer classes
5. **Report as a limitation** - the data may not support trajectory heterogeneity

## Complete Example

```{r complete-example, eval = FALSE}
library(lctmr)
data(sample_growth)

# Step 1: Clean
cleaned <- lctm_clean(
  sample_growth,
  outcome = "weight_raw",
  time_var = "anthroage",
  id_var = "childid"
)

# Step 2: Initial model + diagnostics
initial <- lctm_initial(cleaned, k = 3, degree = 2)
plot(initial, which = "residuals")  # Examine residual patterns

# Step 3: Refine (residuals showed linear trend -> random slope)
result <- lctm_refine(
  initial,
  random = ~ 1 + anthroage,
  k_range = 2:5
)

# Step 4: Interpret
summary(result)
print(result$adequacy)
lctm_plot_trajectories(result$best_model, type = "mean", ci = TRUE)
lctm_plot_trajectories(result$best_model, type = "spaghetti")

# Class assignments for downstream analysis
assignments <- result$class_assignments
head(assignments)
```

## Summary

| Step | Function | Purpose |
|------|----------|---------|
| 1 | `lctm_clean()` | Prepare data, remove implausible values |
| 2 | `lctm_initial()` | Diagnostic plots, guide random effects decision |
| 3 | `lctm_refine()` | Automated BIC search + Model A/B + adequacy |
| 4 | `lctm_plot_*()` | Visualize and interpret results |

The workflow places investigator judgment at the center: you examine plots and residual patterns to decide the trajectory form and random effects structure, then `lctm_refine()` automates the model search.
